---
title: "Data Mining Project - Milkcow Facts"
author: "Nicolas Mora"
date: "9/29/2021"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Introduction to Dataset

Data this week comes from the USDA (United States Department of Agriculture). The raw datasets (Excel Sheets) can be found at - https://www.ers.usda.gov/data-products/dairy-data/documentation/#Loc3

### Loading Libraries

This analysis leverages functions from various libraries, loaded below. Also, the settings for not using scientific notation are also loaded here. 

```{r echo=T, results = 'hide', error=FALSE, warning=FALSE, message=FALSE}

library(tidyverse)
library(scales)
library(dplyr)
library(lubridate)
library(lemon)
library(ggpmisc)
library(ggpubr)
library(devtools)
library(reshape2)
library(plotly)
library(ggplot2)
library(forcats)

options(scipen=999)

```

### Importing Datasets

```{r echo=T, results = 'hide', error=FALSE, warning=FALSE, message=FALSE}

milkcow_facts <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-01-29/milkcow_facts.csv")

fluid_milk_sales <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-01-29/fluid_milk_sales.csv")

milk_product_facts <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-01-29/milk_products_facts.csv")

clean_cheese <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-01-29/clean_cheese.csv")

state_milk_production <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-01-29/state_milk_production.csv")

```

## Initial Exploration of Dataset


```{r}

#Loads the data in a new window for a tabular visualization of the contents. 
View(milkcow_facts)
View(fluid_milk_sales)
View(milk_product_facts)
view(clean_cheese)
view(state_milk_production)

```


**Milkcow Facts** - this dataset shows yearly values from 1980-2014 for various facts that help understand the milk industry such as average price of cows and average milk per cow.

**Fluid Milk Sales** - this dataset shows the sales for different fluid beverage milk products from 1975-2017.

**Milk Product Facts** - this dataset shows the average yearly consumption in lbs per person for various milk products from 1975-2017.

**Clean Cheese** - this dataset shows the average yearly consumption in lbs per person for various cheese products from 1970-2017.

**State Milk Production** - this dataset shows the total milk production values for each state from 1970-2017

### Summary Statistics for Milkcow Facts

```{r}

summary(milkcow_facts)
print(milkcow_facts)

```
+ Year is incorrectly classified as a numeric type, double. This should not cause any problems unless an analysis where order is important is created. On-the-spot class changes will be done in that case.

### Summary Statistics for Fluid Milk Sales

```{r}

summary(fluid_milk_sales)
print(fluid_milk_sales)

```
+ Year is incorrectly classified as a numeric type, double. This should not cause any problems unless an analysis where order is important is created. On-the-spot class changes will be done in that case.

### Summary Statistics for Milk Product Facts

```{r}

summary(milk_product_facts)
print(milk_product_facts)

```
+ Year is incorrectly classified as a numeric type, double. This should not cause any problems unless an analysis where order is important is created. On-the-spot class changes will be done in that case.

### Summary Statistics for Clean Cheese

```{r}

summary(clean_cheese)
print(clean_cheese)

```
+ Data for Blue cheese is missing for a 12 year period, from 1998-2009
+ Year is incorrectly classified as a numeric type, double. This should not cause any problems unless an analysis where order is important is created. On-the-spot class changes will be done in that case.

### Summary Statistics for State Milk Production

```{r}

summary(state_milk_production)
print(state_milk_production)

```

+ Year is incorrectly classified as a numeric type, double. This should not cause any problems unless an analysis where order is important is created. On-the-spot class changes will be done in that case.

## Cleaning the dataset

The datasets all seem to be in good condition for analyses. No facilitating changes are needed. Any changes will be made with the visualizations.


## Visualizing Distributions for "Fluid Milk Sales" dataset

#### What are the popularity trends for fluid milk types?

```{r fig.height=6, fig.width=12}

fluid_milk_sales %>%
  filter(milk_type == "Total Production") %>%
  ggplot(aes(x = pounds)) +
  geom_histogram(bins = 15, show.legend = FALSE) + 
  labs(x = "Pounds", y = "Count" ) + 
  scale_x_continuous()


fluid_milk_sales %>%
  filter(milk_type == "Total Production") %>%
  ggplot(aes(x = year, y = pounds)) +
  geom_line(size = 2) +
  geom_smooth(se = FALSE) +
  scale_x_continuous(breaks = c(1975,1980,1985,1990,1995,2000,2005,2010,2015)) +
  labs(x = "Year", y = "Pounds Sold", title = "Total Pounds of Fluid Milk Sold, 1975-2017") +
  theme_classic()


fluid_milk_sales_graph <- fluid_milk_sales %>%
  group_by(milk_type) %>%
  filter(milk_type != "Total Production") %>%
  ggplot(aes(x = year, y = pounds, color = milk_type)) +
  geom_line(size = 1) +
  scale_x_continuous(breaks = c(1975,1980,1985,1990,1995,2000,2005,2010,2015)) +
  scale_y_continuous(breaks = c(0,5000000000,10000000000,15000000000,20000000000,25000000000,30000000000,35000000000)) +
  labs(x = "Year", y = "Pounds Sold", title = "Pounds of Fluid Milk sold by type, 1975-2017", color = "Milk Type") +
  theme_classic()
ggplotly(fluid_milk_sales_graph)


fluid_milk_sales %>%
  filter(milk_type != "Total Production") %>%
  ggplot(aes(x = pounds, y = milk_type, fill = milk_type)) +
  geom_boxplot(show.legend = FALSE) +
  scale_x_continuous() +
  labs(title = "Variability of Milk Type Sales", x = "Pounds", y = "Milk Type", fill = "Milk Type") +
  theme_gray()




```


+ Overall fluid milk sales have been on a downward trend since 2010, after nearly 25 years (1985-2010) of consistent sales.
+ The demand for whole milk has declined greatly over the years. Meanwhile, 2% reduced fat milk has been on the rise. In the 1980's, whole milk was dominant while 2% reduced fat milk was in low demand. Now, they are very close in terms of demand, with 2% Reduced fat milk being ahead slightly. One can infer that societal views on fats have lead to a decline in Whole milk sales over time.
+ Whole Milk has seen the most variability compared to other fluid types, indicating that, though popular, it is very inconsistent. Skim milk also has relatively high variability while less popular fluid types have been very consistent in their sales throughout the recorded timeframe.
+ The graphic shows us the two most popular liquid milk types and how they have fluctuated in production over the years. Clearly, you could see that the demand for whole milk has declined greatly over the years. Meanwhile, 2% reduced fat milk has been on the rise. In the 1980's, whole milk was dominant while 2% reduced fat milk was in low demand. Now, they are very close in terms of demand, with 2% Reduced fat milk being ahead slightly.


## Visualizing Distributions for "State Milk Production" dataset

#### How is the milk production distributed by location?


```{r fig.height=6, fig.width=12}

state_milk_production %>%
  mutate(state = fct_reorder(state, milk_produced)) %>%
  group_by("state") %>%
  ggplot(aes(x = milk_produced, y = state, fill = region)) +
  geom_boxplot() +
  labs(title = "Variability of Milk Production by State", x = "Milk Production in Lbs", y = "State", fill = "Region") +
  theme_gray()

state_milk_production %>%
  filter(region == "Mountain") %>%
  group_by(state) %>%
  summarise(mean_milk = mean(milk_produced)) %>%
  ggplot(aes(x = state, y = mean_milk, fill = state)) +
  geom_col() +
  labs(title = "Average Milk production by state, Mountain Region",x = "State", y = "Milk Produced", fill = "State") +
  theme_minimal()

state_milk_production %>%
  filter(region == "Lake States") %>%
  group_by(state) %>%
  summarise(mean_milk = mean(milk_produced)) %>%
  ggplot(aes(x = state, y = mean_milk, fill = state)) +
  geom_col() +
  labs(title = "Average Milk production by state, Lake States Region",x = "State", y = "Milk Produced", fill = "State") +
  theme_minimal()

state_milk_production %>% 
  mutate(region = fct_lump(region, n = 5)) %>%
  group_by(region) %>%
  summarise(mean_milk = mean(milk_produced)) %>%
  ggplot(aes(x = region, y = mean_milk, fill = region)) + 
  geom_col() +
  labs(title = "Average Milk production by region", x = "Region", y = "Milk Produced", fill = "Region") +
  theme_minimal()


state_milk_production %>%
  filter(milk_produced >= 10000000000) %>%
  count(state, vars = year) %>%
  ggplot(aes(x = state, fill = state)) +
  geom_bar() +
  labs(title = "# of Years that production exceeded 10B Lbs, by State",x = "State", y = "Count", fill = "State") +
  theme_minimal()


    
```

+ The boxplot represents the variability of milk production by state, color coded by region. We can see that region is not very important for milk production, the only Region that showcases similar milk production values are the Lake States and Mountain states. One can also easily see that California is by far the largest producer, and the only significant producer in its region.
+ The following two graphs show that location and land affect the amount of gallons of milk produced. The states included in the mountain region produce on average twice the amount of gallons of milk that the states included in the lake region. This is because cows live and reproduce better in mountainous areas, therefore leading to a bigger amount of milk produced every year.
+  Tying to the previous graphs, the "Average Milk Production by Year" shows us which region in the US produce more milk. Allowing us to visualize that the Pacific produces more milk than any other region by more than 2 Billion gallons. This might be caused by the temperatures in these stats which allow for a better life for the cows, leading to a bigger amount of milk production.
+ Three of the biggest state producers of milk are California, New York and Wisconsin. For over 45 years these three states have consistently produced over 10 billion gallons of milk. Again, allowing us to conclude that the cow's living conditions is a big factor that affects milk production every year.


## Visualizing Distributions for "Milkcow Facts" dataset

#### What trends are present in the milk production industry?


```{r fig.height=6, fig.width=12}


yrel <- milkcow_facts %>%
  ggplot(aes(x = milk_per_cow, y = avg_milk_cow_number, color = year)) +
  geom_point() +
  geom_smooth(stat = "smooth", method = "lm") +
  labs(title = "Yearly Relationship Between Number of Cows and Milk Per Cow",x = "Milk Per Cow", y = "Average Number of Cows", color = "Year") +
  theme_bw()
ggplotly(yrel)


milkcow_facts %>%
  ggplot(aes(x = milk_per_cow, y = milk_cow_cost_per_animal, color = year)) +
  geom_point(size = 2) +
  geom_line() +
  geom_smooth(se = FALSE, method = "lm") +
  labs(title = "Yearly Relationship Between Milk Per Cow and Cost Per Animal",x = "Milk Per Cow", y = "Average Cost Per Cow", color = "Year") +
  theme_bw()


create_scatter_plot <- function(x,y) {
  ggplot(data = milkcow_facts) +
    aes(x = x, y = y, fill = x, group =1) +
    geom_point(state = "identity") +
    geom_smooth(method = lm, se = FALSE)+
    theme(axis.text.x = element_text(face = "bold", angle = 90)) +
    theme(legend.position = "none")
}

create_scatter_plot(milkcow_facts$avg_price_milk, milkcow_facts$dairy_ration) +
  labs(x= "Average Price of Milk", y = "Average price paid for dairy cow rations", subtitle = "Relationship between Avg. Price of Milk and Avg. Price Paid for Dairy Cow Rations")



```

+ The scatter plot shows the relationship between the milk per cow and number of cows over the years. You can see there is a strong relationship between these two factors over time. Over the years milk per cow has increased, decreasing the number of cows needed. Producers have become better at maintaining the cows for a longer period of time or just producing more milk from the same cow. Allowing them to make higher profits by not having to invest in a higher amount of "inventory".
+ Even though throughout the years producers have been able to produce a higher amount of milk per cow, inflation has had a huge effect on the price of cows on average throughout time. From the 1980 to the year 2010 the average price per cow has increased from less than $1,250 to almost $1,800. 
+ We see that as the price of milk goes up, the price of rations goes up as well. We could assume that this could be due to inflation.


